---
output:
  html_document: default
  pdf_document: default
---
# Chapter 2. Regression and model validation.
Today is:
```{r}
date()
```
#### Task 1. Read and explore the data.
Reading data from local directory.
```{r}
students2014<-read.table(file="data/learning2014.txt",header=T, sep="\t")
```
Success!
Let's check the data structure using **R** functions **dim** and **str**.
```{r}
dim(students2014)
str(students2014)
```
The data has 7 variables from 166 students. 
Variables **deep**, **strategic** and **surface** learning were created based on survey done in 2014. Method to calculate variables is described [here](https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS2-meta.txt). Rest variables are __gender__, __Age__, global __Attitude__ towards statistics, and final exam __Points__. 

#### Task 2. Graphcal overview of the data.
Inbuild **R** function **plot** quite ugly. Therefore I would use [ggplot2](https://ggplot2.tidyverse.org/reference/ggplot.html) as much elegant way to overview the data.

Let's access ggplot package first.
```{r}
library(ggplot2)
```
[GGally](https://ggobi.github.io/ggally/) package will be useful also.
```{r}
library(GGally)
```
Now let's create pairwise plot presenting relationships among variables.

```{r}
ggpairs(students2014, mapping = aes(col=gender,alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)),upper = list(continuous = wrap("cor", size = 2.5)))
```

summary of the data
```{r}
summary(students2014)
```
Number of Female students in the data was larger then Males. The mean students age was 25.51 and the oldest student - 55 years old. Worth to check standard deviation.
```{r}
sd(students2014$Age)
```
The maximum earned exam point was 33, all student's average was 22.72. 
According to ggpairs plot, most of the variables have normal distribution. The students attitude highly correlated with Exam points. No dependency was found between points and gender or age.  
The negative correlation with exam points was observed in deep and surf variables. In opposite strategic learning positively correlated with exam points. 
Let's build linear model...

#### Task 3. Linear model.

Explanatory variables I choose in a first run were attitude, deep, strategic and surface leanings.

The model had look:
```{r}
model=lm(Points~Attitude+stra+deep+surf, students2014)
summary(model)
```
Only Attitude here pass significance test in this model.
The value 1.34e-08 *** in Pr(>|t|) column. 

I have tried different models and find out that only Attitude has a significant relationship with exam points.  

Perfect model is:

```{r}
perfect_model<-lm(Points~Attitude, students2014)
summary(perfect_model)
```
#### Task 4. Explain the summary.
Residuals are not symmetrically distributed.
Alpha parameter is 11.63
Effect of attitude on examine point is 0.352 with standard error 0.056. P test was 4^10-9, which shows significant relationship between target and explanatory variables.
Roughly 19% examine Points variance was explained by Attitude variable, according to R2.

#### Task 5. Diagnostic plots.
Let's create diagnostic plots.
```{r}
plot(perfect_model,which=c(1,2,5))
```

Scatter plot of Residual vs Fitted values presenting random spread of points. 
This means that variance has constant variance. No abnormall pattern was detected.

Normality Q-Q plot presenting distribution of model errors. 
Slight deviation is seen at the beginning and the end of the line.
But generally speaking errors are normally distributed as they following line.

Residuals vs Leverage plot showing absence of outspending observations in the linear model.

All three plots are supporting model linearity assumption and normal distribution of errors.

